---
title: "Exploratory Data Analysis - Beginners R"
author: "Jeremy Ash"
date: "2025-03-25"
output:
  html_document: 
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}

# packages
library(tidyverse)
library(patchwork)

```

# Part A: Load and inspect the data

This notebook analyzes USDA-ERS data on food environment factors (a subset of the Food Environment Atlas), along with county-level food security data from Feeding America, to explore drivers of child poverty rate. We use [this subset of data](https://usdagcc.sharepoint.com/:f:/r/sites/FNS-DataScienceTrainingProgram/Shared%20Documents/FY24-25%20Beginners/Assignments/Assignment%205%20-%20Exploratory%20Data%20Analysis?csf=1&web=1&e=HG79oc) to create summary statistics, scatter plots, histograms, and correlations.


## A.1: Loading the data

First, we import the csv file into a data frame object called `food_env`. 


```{r part_a_1, warning=FALSE, message = FALSE}


# read data into R
# note: the working directory here is the location of the .Rmd file, so I have put the csv file in the same location
food_env <- read_csv("EDA_assignment_dataset.csv")


```
## A.2: Summary statistics

Next, we explore the basic summary statistics for each column. A quick way to do this is to use the `summary()` function. If we were interested in particular summary statistics, we can also create a new dataframe with summary measures using `tidyverse` functions. 

```{r part_a_2, warning=FALSE, messages = FALSE}

# basic summary stats of variables in the food_env dataset, showing the number of NAs per variable
summary(food_env)

# custom set of summary stats for CHILDPOVRATE15
childpovrate15_summary <- food_env %>% 
  summarise(n = length(na.omit(CHILDPOVRATE15)), # number of observations with non-NA values
            n_na = length(which(is.na(CHILDPOVRATE15))), # number of NA values
            min = min(CHILDPOVRATE15, na.rm = TRUE), # min value, excluding NA
            max = max(CHILDPOVRATE15, na.rm = TRUE), # max value, excluding NA
            mean = mean(CHILDPOVRATE15, na.rm = TRUE), # mean value, excluding NA
            median = median(CHILDPOVRATE15, na.rm = TRUE), # median value, excluding NA
  )
  
childpovrate15_summary

```

## A.3: Exclude NA values

We see above that there are various amounts of `NA` values in each column. To continue with an analysis focused only on complete cases, we next create a dataset that excludes any rows with `NA` values.

```{r part_a_3, warning=FALSE, message = FALSE}

# create a new data set with only complete observations
food_env_full <- food_env %>% 
  drop_na() # exclude any row containing NA values

# create a variable showing the length of food_env
food_env_n <- dim(food_env)[1]
food_env_n

# create a variable showing the length of food_env_full to see how many NA cases were dropped
food_env_full_n <- dim(food_env_full)[1]
food_env_full_n
```

Above, we can see the difference in the number of rows of each dataframe is 29. In case this hasn't been shown before, you can also use inline code in `Rmd` files (look at `Rmd` file to see how this looks in code), so we could say that in the original dataset, there were `r food_env_n` observations. After excluding `NA` values, there are `r food_env_full_n` observations. 


# Part B: Explore child poverty rates

Now we want to start visualizing the data, specifically the child poverty rates from 2015. Below is a histogram showing the distribution of child poverty rates from 2015, with vertical lines indicating the mean and median values for this particular variable. We see that the median is lower than the mean, given the skewness of the data.

```{r part_b, warning=FALSE, message = FALSE}
# note: we have already calculated the mean and median values of CHILDPOVRATE15 in childpovrate15_summary, but first we will reshape that data into a simplified dataframe 

childpovrate15_lines <- childpovrate15_summary %>% 
  
  # select the two summary measures of interest
  select(mean, median) %>% 
  
  # gather into a long format
  gather(key = "stat", value = "val")
  


# using ggplot2, we can create the histogram
childpovrate15_hist <- ggplot() +
  
  # create histogram of values
  geom_histogram(aes(x = CHILDPOVRATE15), 
                 data = food_env_full) + 
  
  # plot mean and median values, with different linetypes
  geom_vline(aes(xintercept = val, linetype = stat), 
             key_glyph = "path", # this changes legend lines to be horizontal
             data = childpovrate15_lines) +
  
  # set axis limits for later figure comparisons
  xlim(0, childpovrate15_summary$max+5) +
  
  # use a minimal theme 
  theme_minimal() +
  
  # add a title
  labs(title = "Distribution of National Child Poverty Rates, 2015")


# display the figure
childpovrate15_hist


```



# Part C: Explore low access to stores and no vehicle access

Before we investigate correlative relationships, let's look at the distribution of one potential factor in child poverty rates: low accessibility to stores and no vehicle access. 

## C.1: Basic distribution

Below is a histogram and boxplot showing the overall distribution of the variable `PCT_LACCESS_HHNV15`. We can see in the histogram a long tail to the right, showing that there are some counties with a high percentage of households with no car and low access to stores. This is confirmed looking at the boxplot, where we see a number of points falling to the right of the "upper whisker". 

```{r part_c_1, warning=FALSE, message = FALSE}

# lets create a dataframe where we can again plot the mean and median values
pct_lacess_hhnv15_lines <- food_env_full %>% 
  
  # summarise using food_env_full, so no NA values present
  summarise(mean = mean(PCT_LACCESS_HHNV15), # mean value
            median = median(PCT_LACCESS_HHNV15) # median value
  ) %>% 
  
  # gather into a long format
  gather(key = "stat", value = "val")


# using ggplot2, we can create the histogram
pct_lacess_hhnv15_hist <- ggplot() +
 
   # create histogram of values
  geom_histogram(aes(x = PCT_LACCESS_HHNV15), 
                 data = food_env_full) + 
  
  # plot mean and median values, with different linetypes
  geom_vline(aes(xintercept = val, linetype = stat), 
             key_glyph = "path", # this changes legend lines to be horizontal
             data = pct_lacess_hhnv15_lines) +
  
  # use a minimal theme 
  theme_minimal() +
  
  # add a title
  labs(title = "Distribution of Percentage of Households with No Vehicle/Low Store Access")


# display the histogram
pct_lacess_hhnv15_hist


# similarly, using ggplot2, we can create the boxplot
pct_lacess_hhnv15_boxplot <- ggplot() +
  
  # create boxplot of values
  geom_boxplot(aes(x = PCT_LACCESS_HHNV15), 
                 data = food_env_full) + 
  
  # use a minimal theme 
  theme_minimal() +
  
  # add a title
  labs(title = "Distribution of Percentage of Households with No Vehicle/Low Store Access")


# display the histogram
pct_lacess_hhnv15_boxplot

```


## C.2: Access by metro/non-metro status

Adding more detail to our exploratory analysis, we will now look at whether the distribution of `PCT_LACCESS_HHNV15` differs between metro and non-metro areas. It's not obvious to me what the differences are by looking at the histogram, but the differences are clearer looking at the boxplots. Here we can see that households in non-metro areas are more likely to have no vehicle and lower access to stores, as you might expect.

```{r part_c_2, warning=FALSE, message = FALSE}

# first, lets use Metro verus Non-metro instead of 1 and 0, respectively
metro_labeller <- c(
  `1` = "Metro",
  `0` = "Non-metro"
)

# using ggplot2, we can create the histogram
pct_lacess_metro_hist <- ggplot() +
  
  # create histogram of values 
  geom_histogram(aes(x = PCT_LACCESS_HHNV15), 
                 data = food_env_full) + 
  
  # split out by metro status
  facet_grid(rows = vars(METRO13),
             # use labeller object created above
             labeller = as_labeller(metro_labeller)) +
 
  
  # use a minimal theme 
  theme_minimal() +
  
  # add a title
  labs(title = "Distribution of Percentage of Households with No Vehicle/Low Store Access")

# display the faceted histogram
pct_lacess_metro_hist


# using ggplot2, we can create the faceted boxplot
pct_lacess_metro_boxplot <- ggplot() +
  
  # create boxplot of values 
  geom_boxplot(aes(x = PCT_LACCESS_HHNV15), 
                 data = food_env_full) + 
  
  # split out by metro status
  facet_grid(rows = vars(METRO13),
             # use labeller object created above
             labeller = as_labeller(metro_labeller)) +
 
  
  # use a minimal theme 
  theme_minimal() +
  
  # add a title
  labs(title = "Distribution of Percentage of Households with No Vehicle/Low Store Access")

# display the faceted boxplot
pct_lacess_metro_boxplot

```



# Part D: Exploring relationships

Now that we have investigated the distribution of individual variables, we'll get to the more question-driven portion of the analyses and explore whether variables correlate with one another. For this section, we will look to see if there is a relationship between `PCT_LACCESS_HHNV15` and `CHILDPOVRATE15`. In other words, does access to vehicles and stores affect the child poverty rate in counties across the country?


## D.1: Correlation and scatterplot
To address this question, we will look at the statistical correlation between these variables, as well as a scatterplot to visualize the relationship. We will calculate the Pearson's correlation coefficient to measure the linear correlation between `PCT_LACCESS_HHNV15` and `CHILDPOVRATE15`. We will also show the relationship visually using a scatterplot.


```{r part_d_1, warning=FALSE, message = FALSE}

# calculate the correlation by specifying the two variables and Pearson as function inputs
correlation <- cor(food_env_full$PCT_LACCESS_HHNV15,
                   food_env_full$CHILDPOVRATE15,
                   method = 'pearson')

correlation

# generate and view the scatterplot
access_chpov_scatter <- ggplot() +
  
  # generate the points with access on the x-axis and child poverty on the y
  geom_point(aes(x = PCT_LACCESS_HHNV15,
                 y = CHILDPOVRATE15),
             data = food_env_full) +
  
  # use a simple theme
  theme_minimal() +
  
  # add a title
  labs(title = "Relationship between Child Poverty and No Vehicle/Low Store Access")
  
access_chpov_scatter

```


## D.2: Interpretation of initial results

From the scatterplot, we can see evidence of some relationship between the percent of households with no vehicle and low store access (higher values on the x axis) and child poverty rates. That being said, there is a lot of variability in the data (focusing here on the data with `PCT_LACCESS_HHNV15 < 20`) and counties with high percentages of household without vehicles and store access (`PCT_LACCESS_HHNV15 > 20`) experience moderate rates of child poverty. The Pearson's correlation coefficient (`r round(correlation, 3)`) confirms this, as it shows a moderate, positive linear relationship between the two variables. It's clear, however, that the relationship is not strongly predictive. One other factor not captured in this dataset that might explain more of the variability is access and availability of public transit in the county. 



## D.3: Influence of metro status on the relationship

As before, we will now explore whether we can observe a difference when we consider counties within metro versus non-metro areas.

```{r part_d_3, warning=FALSE, message = FALSE}

# to start, create separate dataframes of metro and non-metro data
metro_df <- food_env_full %>% 
  
  # filter to metro areas
  filter(METRO13 == 1)

non_metro_df <- food_env_full %>% 
  
  # filter to non-metro areas
  filter(METRO13 == 0)


# calculate the correlation by specifying the two variables and Pearson as function inputs by metro df
metro_correlation <- cor(metro_df$PCT_LACCESS_HHNV15,
                         metro_df$CHILDPOVRATE15,
                         method = 'pearson')

metro_correlation

# calculate the correlation by specifying the two variables and Pearson as function inputs by metro df
non_metro_correlation <- cor(non_metro_df$PCT_LACCESS_HHNV15,
                             non_metro_df$CHILDPOVRATE15,
                             method = 'pearson')

non_metro_correlation

# generate and view the scatterplot using the full dataset and the previously generated labeller
metro_access_chpov_scatter <- ggplot() +
  
  # generate the points with access on the x-axis and child poverty on the y
  geom_point(aes(x = PCT_LACCESS_HHNV15,
                 y = CHILDPOVRATE15),
             data = food_env_full) +
  
  # split out by metro status
  facet_grid(rows = vars(METRO13),
             # use labeller object created above
             labeller = as_labeller(metro_labeller)) +
  
  
  # use a simple theme
  theme_minimal() +
  
  # add boxes around panels so they are more easily distinguishable
  theme(panel.border = element_rect(linewidth = 2, fill = "transparent")) +
  
  # add a title
  labs(title = "Relationship between Child Poverty and No Vehicle/Low Store Access")
  
metro_access_chpov_scatter

```

While the relationship remains variable for metro areas (Pearson's correlation coefficient  = `r round(metro_correlation, 3)`), there does appear to be a stronger correlation than in non-metro areas (Pearson's correlation coefficient  = `r round(non_metro_correlation, 3)`). 


# BONUS

I live in North Carolina, so I'm interested in exploring child poverty rate patterns across the state. The histogram and mean/median values show that child poverty rates in NC are higher than the national averages. A suite of lacking social support systems, cost of living, wages, etc. are likely contributors.  

```{r bonus_1, warning=FALSE, message = FALSE}

# to start, create separate dataframe for NC
nc_df <- food_env_full %>% 
  
  # filter to metro areas
  filter(State == "NC")


# lets create a dataframe where we can again plot the mean and median values
nc_childpovrate15_lines <- nc_df %>% 
  
  # summarise using food_env_full, so no NA values present
  summarise(mean = mean(CHILDPOVRATE15), # mean value
            median = median(CHILDPOVRATE15) # median value
  ) %>% 
  
  # gather into a long format
  gather(key = "stat", value = "val")

# I want to be able to compare directly to the national figure so I am going to set the x axis limits on this figure to be the same as the national figure

# using ggplot2, we can create the histogram
nc_childpovrate15_hist <- ggplot() +
  # create histogram of values
  geom_histogram(aes(x = CHILDPOVRATE15), 
                 data = nc_df) + 
  
  # plot mean and median values, with different linetypes
  geom_vline(aes(xintercept = val, linetype = stat), 
             key_glyph = "path", # this changes legend lines to be horizontal
             data = nc_childpovrate15_lines) +
  
  # same limits as national figure
  xlim(0, childpovrate15_summary$max+5) +
  
  
  # use a minimal theme 
  theme_minimal() +
  
  # add a title
  labs(title = "Distribution of NC Child Poverty Rates, 2015")


# display the figure atop the national figure using the patchwork library
nc_childpovrate15_hist/childpovrate15_hist



```


Below is the code to explore the correlation for counties within NC and the scatterplot with a linear regression line. 

```{r bonus_2, warning=FALSE, message = FALSE}
# calculate the correlation by specifying the two variables and Pearson as function inputs
nc_correlation <- cor(nc_df$PCT_LACCESS_HHNV15,
                      nc_df$CHILDPOVRATE15,
                      method = 'pearson')

nc_correlation



# generate and view the scatterplot using the full dataset and the previously generated labeller
nc_access_chpov_scatter <- ggplot() +
  
  # generate the points with access on the x-axis and child poverty on the y
  geom_point(aes(x = PCT_LACCESS_HHNV15,
                 y = CHILDPOVRATE15),
             data = nc_df) +
  
  # add the regression line
  geom_smooth(aes(x = PCT_LACCESS_HHNV15,
                  y = CHILDPOVRATE15),
              method='lm', 
              data = nc_df) +
  
  # use a simple theme
  theme_minimal() +
  
  # add boxes around panels so they are more easily distinguishable
  theme(panel.border = element_rect(linewidth = 2, fill = "transparent")) +
  
  # add a title
  labs(title = "Relationship between NC Child Poverty and No Vehicle/Low Store Access")
  
nc_access_chpov_scatter

```

Looking at the correlation between `PCT_LACCESS_HHNV15` and `CHILDPOVRATE15`, we see a stronger relationship between these variables than in the national analyses (Pearson's correlation coefficient for NC = `r round(nc_correlation, 3)`). While again, there is variability in the scatterplot, the positive relationship betweent these two variables is more apparent. The regression line fit to these data shows this positive relationship. 




